---
title: "Microglia GSEA across brain"
subtitle: "Aducanumab vs IgG, per-region fgsea against MSigDB M7 (Immunologic Signatures) mouse gene sets"
date: last-modified
---

```{r setup}
#| message: false
library(msigdbr)
library(fgsea)
library(tidyverse)
library(kableExtra)
library(ggrepel)
library(pheatmap)
```

# Overview

Gene Set Enrichment Analysis (GSEA) was applied to characterise the immunological pathways
altered in microglia by aducanumab treatment, across individual brain regions. The analysis
uses ranked gene lists derived from the pseudobulk DESeq2 results using the _clean Kai´s microglia seurat object_ and tests them against the
[M7 Immunologic Signature](https://www.gsea-msigdb.org/gsea/msigdb/mouse/collection_details.jsp#M7)
gene sets from MSigDB.

**M7 rationale.** M7 gene sets represent cell states and perturbations within the immune
system, with the initial release curated from
[Cui et al. 2023](https://www.nature.com/articles/s41586-023-06816-9). These sets catalogue
cell-type-specific transcriptional responses of mouse immune cells to 86 cytokines *in vivo*,
making them highly appropriate for interpreting microglial responses to immunotherapy. See also the [Immune Dictionary Portal](https://immune-dictionary.org/app/home) for more information.

**Panel constraint.** The Xenium panel targets approximately 480 genes. Because GSEA only ranks
genes present in the data, the effective size of each gene set reflects its intersection with
this panel. To retain gene sets with as few as five panel-represented members, `minSize = 5`
was used (compared with the conventional threshold of 15 used for whole-transcriptome data).
No explicit panel filter was applied: `fgsea` automatically intersects the ranked list with each
pathway, so effective sizes already reflect the panel constraint.

**Interpretation of NES.** The Normalized Enrichment Score (NES) measures whether a gene set
is over-represented at the top or bottom of the ranked gene list. It is normalized for gene set
size, allowing comparison across pathways and regions. NES > 0 indicates that genes in the set
are disproportionately upregulated in aducanumab-treated microglia; NES < 0 indicates
disproportionate upregulation in IgG controls. Pathway redundancy within each region was
reduced using `collapsePathways` (threshold *p* < 0.01) before cross-region comparison.

---

# Setup

## Load M7 gene sets

```{r load-gene-sets}
gene_sets <- msigdbr(species = "Mus musculus", db_species = "MM", collection = "M7")
gene_sets_list <- split(gene_sets$gene_symbol, gene_sets$gs_name)

cat("M7 gene sets loaded:", length(gene_sets_list), "\n")
```

## Load DE results

```{r load-de-results}
de_by_region <- readRDS("results/20260303-DEG_MG_byRegion.rds")

# Panel genes = union of all genes tested across regions
panel_genes <- de_by_region |> map(\(df) rownames(df)) |> reduce(union)

cat("Regions available:", paste(names(de_by_region), collapse = ", "), "\n")
cat("Panel genes:", length(panel_genes), "\n")
```

## Panel coverage in M7

The table below quantifies how many panel genes are represented in the M7 collection,
and how many M7 gene sets have at least one panel member.

```{r panel-coverage}
m7_genes <- unique(gene_sets$gene_symbol)

tibble(
  panel_size            = length(panel_genes),
  m7_unique_genes       = length(m7_genes),
  panel_in_m7           = sum(panel_genes %in% m7_genes),
  pct_covered           = round(100 * sum(panel_genes %in% m7_genes) / length(panel_genes), 1),
  genesets_with_overlap = sum(map_int(gene_sets_list, \(g) sum(g %in% panel_genes)) > 0)
) |>
  kbl(caption = "Panel coverage in M7 gene sets") |>
  kable_styling("striped")
```

---

# Per-region GSEA

For each brain region, genes are ranked by `avg_log2FC` (Adu vs IgG), and `fgsea` is run
against all M7 gene sets. Results are displayed as a volcano plot of NES vs
−log~10~(adjusted *p*). All points are plotted; significant pathways (FDR < 0.05) are
highlighted in red.

Because M7 contains many gene sets with overlapping membership, redundant significant pathways
are reduced using `collapsePathways` (fgsea; *p*-value threshold 0.01). This procedure
identifies a subset of *representative* (non-redundant) pathways (*collapsed pathways*), such that no two retained pathways share an unexpectedly large fraction of their
leading-edge genes.  

For clarity, only these collapsed pathways are labelled on the volcano;
the remaining significant-but-redundant pathways remain visible as unlabelled red points.
Pathway labels are cleaned by stripping the leading collection prefix (e.g. `GSE12345_`,
`GOLDRATH_`) and converting underscores to spaces.

```{r per-region-gsea}
#| results: asis
#| fig.width: 10
#| fig.height: 7

gsea_by_region <- imap(de_by_region, \(de_res, reg) {

  # Build ranked list: avg_log2FC sorted descending, all panel genes
  gene_ranks <- de_res |>
    rownames_to_column("gene") |>
    arrange(desc(avg_log2FC)) |>
    (\(d) setNames(d$avg_log2FC, d$gene))()

  res <- fgsea(
    pathways = gene_sets_list,
    stats    = gene_ranks,
    eps      = 0.0,
    minSize  = 5,    # lowered from 15: panel has only ~480 genes
    maxSize  = 500
  )

  sig <- res |> filter(padj < 0.05) |> arrange(desc(NES))

  cat("\n\n##", reg, "\n\n")

  if (nrow(sig) == 0) {
    cat("_No significant pathways (padj < 0.05)._\n\n")
    return(list(fgsea = res, significant = sig, gene_ranks = gene_ranks))
  }

  # Collapse redundant pathways within region
  collapsed <- collapsePathways(
    fgseaRes       = res |> filter(padj < 0.05),
    pathways       = gene_sets_list[sig$pathway],
    pval.threshold = 0.01,
    stats          = gene_ranks
  )
  sig_collapsed <- sig |> filter(pathway %in% collapsed$mainPathways)

  # NES volcano plot
  volcano_data <- res |>
    mutate(
      neg_log10_padj = -log10(padj),
      significant    = padj < 0.05,
      pathway_label  = pathway |>
        str_remove("^[A-Z0-9]+_") |>
        str_replace_all("_", " ") |>
        str_to_sentence()
    )

  p <- ggplot(volcano_data, aes(x = NES, y = neg_log10_padj)) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(aes(size = size, colour = significant), alpha = 0.6) +
    geom_text_repel(
      data = \(d) filter(d, pathway %in% collapsed$mainPathways),
      aes(label = pathway_label),
      size = 6, max.overlaps = 20, segment.colour = "grey60"
    ) +
    scale_colour_manual(
      values = c("TRUE" = "#E64B35", "FALSE" = "grey70"),
      labels = c("TRUE" = "FDR < 0.05", "FALSE" = "NS")
    ) +
    scale_size_continuous(range = c(1, 6), name = "Gene set size") +
    labs(
      title  = paste("GSEA M7 \u2014", reg),
      x      = "Normalized Enrichment Score (NES)",
      y      = expression(-log[10](p[adj])),
      colour = "Significance"
    ) +
    theme_bw(base_size = 16) +
    theme(panel.grid = element_blank(), legend.position = "bottom")
  print(p)

  list(fgsea = res, significant = sig_collapsed, gene_ranks = gene_ranks)

})
```

---

# Cross-region NES heatmap

The heatmap displays only **collapsed pathways**. Taking the
union of each region's collapsed main pathways yields a de-duplicated catalogue of biologically
distinct immunological programmes altered by aducanumab, free of the inflation that arises when
many highly overlapping gene sets reach significance simultaneously.

For each retained pathway, the NES from every region is assembled into a matrix and displayed
as a hierarchically clustered heatmap. Asterisks mark cells where the pathway also meets
FDR < 0.05 in that region. 

```{r nes-heatmap}
#| fig.width: 12
#| fig.height: 10

all_gsea_long <- imap(gsea_by_region, \(res, reg) {
  res$fgsea |> mutate(Region = reg)
}) |> list_rbind()

# Union of collapsed (non-redundant) main pathways across all regions
sig_pathways <- gsea_by_region |>
  map(\(res) res$significant$pathway) |>
  reduce(union)

if (length(sig_pathways) > 0) {
  clean_label <- function(x) {
    x |>
      str_remove("^[A-Z0-9]+_") |>
      str_replace_all("_", " ") |>
      str_to_sentence()
  }

  nes_matrix <- all_gsea_long |>
    filter(pathway %in% sig_pathways) |>
    pivot_wider(id_cols = pathway, names_from = Region, values_from = NES) |>
    mutate(pathway = clean_label(pathway)) |>
    column_to_rownames("pathway") |>
    as.matrix()

  padj_matrix <- all_gsea_long |>
    filter(pathway %in% sig_pathways) |>
    pivot_wider(id_cols = pathway, names_from = Region, values_from = padj) |>
    mutate(pathway = clean_label(pathway)) |>
    column_to_rownames("pathway") |>
    as.matrix()

  display_matrix <- ifelse(!is.na(padj_matrix) & padj_matrix < 0.05, "*", "")
  clim <- ceiling(max(abs(nes_matrix), na.rm = TRUE))

  pheatmap(nes_matrix,
    color           = colorRampPalette(c("#4575b4", "white", "#d73027"))(100),
    breaks          = seq(-clim, clim, length.out = 101),
    na_col          = "grey88",
    display_numbers = display_matrix,
    fontsize_number = 24,
    number_color    = "black",
    cluster_rows    = TRUE,
    cluster_cols    = TRUE,
    main            = "M7 GSEA NES (Adu vs IgG) by region\n",
    fontsize        = 15,
    fontsize_row    = 14,
    angle_col       = 45,
    cellwidth       = 30,
    cellheight      = 30
  )
} else {
  message("No significant pathways across any region — heatmap skipped.")
}
```

---

# Leading-edge tables

Collapsible tables listing significant (after `collapsePathways` within each region) M7 pathways
per region, with NES, adjusted *p*-value, effective gene set size, and the leading-edge genes
driving enrichment.

```{r leading-edge-tables}
#| results: asis

iwalk(gsea_by_region, \(res, reg) {
  cat("\n\n<details><summary>**", reg, "** significant pathways</summary>\n\n")

  if (nrow(res$significant) > 0) {
    res$significant |>
      select(pathway, NES, padj, size, leadingEdge) |>
      mutate(
        pathway     = str_remove(pathway, "^[A-Z0-9]+_") |>
                      str_replace_all("_", " ") |>
                      str_to_sentence(),
        padj        = signif(padj, 3),
        NES         = round(NES, 2),
        leadingEdge = map_chr(leadingEdge, \(x) paste(x, collapse = ", "))
      ) |>
      arrange(padj) |>
      kbl(caption = paste("M7 GSEA \u2014", reg)) |>
      kable_styling("striped") |>
      print()
  } else {
    cat("No significant pathways.\n\n")
  }

  cat("\n\n</details>\n\n")
})
```
