---
title: "Astrocyte subpopulations"
subtitle: "Identification of Astrocyte subpopulations after subsetting and reclustering"
date: last-modified
format:
  html:
    number-sections: false
---

```{r}
#| label: setup
#| message: false
library(qs2)
library(Seurat)
library(patchwork)
library(tidyverse)
library(kableExtra)
library(DESeq2)
library(gridExtra)


astro <- qs2::qs_read("seurat_objects/20260305-astro_0.3.qs2")

```


```{r}
#| eval: false
#| label: SCTransform
# Load full object 
full    <- qs_read("seurat_objects/20260217_Part2_fullobj.qs2")
astro0  <- subset(full, subset = annotatedclusters == "Astrocyte")
# Keep only the assay needed for fresh SCTransform
DefaultAssay(astro0) <- "Xenium"
astro0 <- DietSeurat(astro0, assays = "Xenium", dimreducs = NULL, graphs = NULL)
# Drop Xenium image/FOV segmentation data; otherwise merge triggers huge future globals
astro0@images <- list()

# SCTransform on all astrocytes together (no split/merge, no integration)
astro <- astro0
astro <- SCTransform(
  astro,
  assay = "Xenium",
  new.assay.name = "SCT",
  verbose = FALSE
)
```




## PCA and dimensionality selection

```{r}
#| eval: false
#| label: Run PCA on SCT assay
DefaultAssay(astro) <- "SCT"
astro <- RunPCA(
  astro,
  assay   = "SCT",
  features = VariableFeatures(astro),
  verbose = FALSE)
```
```{r}
#| fig-width: 5
#| fig-height: 3
#| label: Elbow plot
# Elbow plot to visualise variance explained per PC
ElbowPlot(astro, ndims = 50) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(title = "Elbow plot") +
  theme_minimal()
```

```{r}
#| fig-height: 5
#| label: PCA biplots

# PCA biplots coloured by sample and treatment
p1 <- DimPlot(astro, reduction = "pca", group.by = "Slide") +
  labs(title = "PCA — Slide")
p2 <- DimPlot(astro, reduction = "pca", group.by = "Treatment.Group") +
  labs(title = "PCA — Treatment")
p1 + p2
```

```{r}
#| label: neighbors & UMAP
#| eval: false

# Set seed for reproducibility
set.seed(1234)

# Compute nearest neighbors using first 30 PCs
astro <- FindNeighbors(astro, reduction = "pca", dims = 1:30, graph.name = "astro_snn")

# Perform clustering at multiple resolutions for later comparison
astro <- FindClusters(astro, graph.name = "astro_snn", resolution = 0.3)

# Compute UMAP for visualization
astro <- RunUMAP(astro, reduction = "pca", dims = 1:30, verbose = FALSE)

# qs_save(astro, "seurat_objects/20260305-astro_0.3.qs2")
```

resolution = 0.3 

```{r}
#|fig-width: 5
#|fig-height: 5
#| label: plot UMAPs 1
DimPlot(astro)
```

```{r}
#| fig-height: 5
#| label: plot UMAPs 2
DimPlot(astro, split.by = "Treatment.Group")
```

## Top marker genes per cluster

`FindAllMArkers` run on log-normalized residual expression values from SCTransform in the SCT assay (Values are continuous and typically centered around 0)
```{r}
#| eval: false
#| label: find markers
Idents(astro) <- astro$seurat_clusters
DefaultAssay(astro) <- "SCT"
all_markers <- FindAllMarkers(astro, only.pos = TRUE, densify = TRUE)
# write_csv(all_markers, file = "results/20260305-astrocyte_all_markers_0.3.csv")
```

```{r}
#| label: top markers per cluster
#| column: screen
#| fig-width: 15
#| fig-height: 4
source("20260217-helper_functions.R")

# Top 5 markers per cluster by avg_log2FC
top_genes <- read_csv("results/20260305-astrocyte_all_markers_0.3.csv",
                      show_col_types = FALSE) |>
  group_by(cluster) |>
  arrange(desc(avg_log2FC), .by_group = TRUE) |>
  slice_head(n = 5) |>
  pull(gene) |>
  unique()

fiddle(astro,
       genes_of_interest = top_genes,
       classification_col = "seurat_clusters",
       assay = "SCT")
```

### Cluster identity assessment

Based on the top 5 marker genes per cluster, we assessed whether each cluster represents true astrocytes or contaminating cell types from the initial subsetting.

| Cluster | Top markers | Identity | Astrocyte? |
|---------|------------|----------|------------|
| 0 | Agt, Calb2, Apoc1, Pgm1, Folh1 | Astrocyte — Agt is a canonical astrocyte marker | Yes |
| 1 | Fezf2, Hsd11b1, Vim, Plpp3, Scg3 | Astrocyte — Vim and Hsd11b1 are astrocyte-enriched | Yes |
| 2 | Lamp5, Slc17a7, Egr1, Grin2a, Vip | Neurons — Slc17a7 (Vglut1) is glutamatergic; Lamp5/Vip are interneuron markers | No |
| 3 | Cd209a, Mgl2, Kcnj8, Slc47a1, Mrc1 | Perivascular macrophages / pericytes — Mrc1 (CD206), Cd209a, Mgl2 are macrophage markers; Kcnj8 is pericyte | No |
| 4 | Nlrp3, Cst7, Cd14, Itgax, Trem2 | Microglia (DAM) — classic disease-associated microglia signature | No |
| 5 | Mog, Mag, Cldn11, Ermn, Aspa | Oligodendrocytes — all canonical oligodendrocyte markers | No |
| 6 | Ido1, Nts, Tac1, Otof, Slc17a6 | Neurons — Slc17a6 (Vglut2) and Tac1 are neuronal markers | No |
| 7 | Cxcl10, Acod1, Ccl2, Ccl12, Tnf | Reactive astrocytes or immune — Cxcl10/Ccl2 can mark reactive astrocytes, but Tnf/Acod1 suggest immune contamination | Uncertain |
| 8 | Cldn5, Emcn, Flt1, Cxcl12, Klf4 | Endothelial cells — Cldn5, Flt1, Emcn are canonical endothelial markers | No |
| 9 | Ttr, Tmem212, Ccdc153, Lypd1, Otof | Ependymal / choroid plexus — Ttr, Ccdc153, Tmem212 are ependymal markers | No |
| 10 | Pdgfra, Tnr, Pllp, Plpp4, Lpcat2 | OPCs — Pdgfra is the canonical OPC marker | No |

## Astrocyte markers


```{r}
#| label: astrocyte-markers-featureplot
#| column: page
#| fig-height: 12
#| fig-width: 10

FeaturePlot(astro,features = c("Aldh1l1",	"Aldoc",	"Gfap",	"Slc7a10",	"Aqp4"), ncol=2, min.cutoff = "q20")
```

```{r}
#| label: astrocyte-markers-violin
#| column: page
#| fig-height: 12
#| fig-width: 10

VlnPlot(astro,features = c("Aldh1l1",	"Aldoc",	"Gfap",	"Slc7a10",	"Aqp4"),alpha = 0, ncol=2)
```

## Purity score

```{r}
#| column: page
#| fig-height: 5
#| fig-width: 10
#| label: split-UMAP

Idents(astro) <- "first_type_class"
sp <- DimPlot(astro, label =T, repel = TRUE)+ NoLegend()+ ggtitle("SPLIT ???")
Idents(astro) <- "seurat_clusters"
sc <- DimPlot(astro, label =T, repel = TRUE)+ NoLegend()

sp+sc
```


```{r}
FeaturePlot(astro,features = "singlet_score",min.cutoff = "q25")
```
## Differential expression by treatment

Pseudobulk DESeq2 was performed on each true astrocyte cluster (0 and 1) separately,
comparing aducanumab (Adu) vs isotype control (IgG) pooled across all brain regions.
Raw Xenium counts were aggregated per biological sample using `AggregateExpression`,
yielding one pseudobulk profile per brain (n = 3 Adu, n = 3 IgG per cluster).
Significance threshold: adjusted p-value < 0.1.

```{r pseudobulk-de-clusters}
#| eval: false
#| message: false
#| warning: false
#| label: pseudobulk-de-clusters

source("20260217-helper_functions.R")

# Run pseudobulk DESeq2 for each true astrocyte cluster
DefaultAssay(astro) <- "Xenium"
clusters_of_interest <- c(0, 1, 7)

de_by_cluster <- map(clusters_of_interest, \(cl) {
  # Subset to cells in this cluster
  sub_obj <- subset(astro, seurat_clusters == cl)

  # Pseudobulk aggregation per sample
  bulk <- AggregateExpression(sub_obj, group.by = "sample_ID", return.seurat = TRUE)

  # Map treatment metadata (AggregateExpression converts _ to -)
  sample_meta <- distinct(sub_obj@meta.data, sample_ID, Treatment.Group) |>
    mutate(bulk_id = gsub("_", "-", sample_ID))
  bulk$Treatment.Group <- sample_meta$Treatment.Group[
    match(colnames(bulk), sample_meta$bulk_id)
  ]
  Idents(bulk) <- "Treatment.Group"

  # DESeq2 via anayet helper
  # Relaxed threshold (p_adj < 0.1) for astrocyte subclusters
  anayet(bulk,
         ident1          = "Adu",
         ident2          = "IgG",
         test_method     = "DESeq2",
         min_pct         = 0,
         p_adj_threshold = 0.1,
         plot_title      = paste("Astrocyte cluster", cl, "— Adu vs IgG"),
         print_plot      = FALSE)

}) |> set_names(paste0("cluster_", clusters_of_interest))

# saveRDS(de_by_cluster, "results/20260305-DEG_astrocytes_clusters0-1.rds")
```


```{r display-de-clusters}
#| results: asis
#| column: page
#| fig.width: 14
#| fig.height: 6
#| warning: false
#| label: display-de-clusters
de_by_cluster <- readRDS("results/20260305-DEG_astrocytes_clusters0-1.rds")

# Display volcano plot + significant genes table for each cluster
walk(names(de_by_cluster), \(cl) {
  res <- de_by_cluster[[cl]]
  label <- gsub("_", " ", cl) |> str_to_title()
  cat("\n\n###", label, "\n\n")

  sig_tbl <- if (nrow(res$significant_results) > 0) {
    res$significant_results |>
      select(p_val, avg_log2FC, p_val_adj) |>
      mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
      arrange(p_val_adj) |>
      tableGrob(theme = ttheme_minimal(base_size = 18))
  } else {
    tableGrob(
      data.frame(Result = "No significant DEGs"),
      theme = ttheme_minimal(base_size = 20)
    )
  }

  print(res$volcano_plot + wrap_elements(sig_tbl) + plot_layout(widths = c(2, 1)))
})
```

