---
title: "Astrocyte cluster characterization"
subtitle: "Characterization of astrocyte clusters 0, 1, and 7 from reclustering at resolution 0.3"
date: last-modified
format:
  html:
    number-sections: false
---

```{r setup}
#| message: false
library(qs2)
library(Seurat)
library(patchwork)
library(tidyverse)
library(kableExtra)
library(DESeq2)
library(gridExtra)
library(grid)
library(compareGroups)
library(pheatmap)

source("20260217-helper_functions.R")

# Load reclustered astrocyte object (resolution 0.3, 11 clusters)
astro <- qs_read("seurat_objects/20260305-astro_0.3.qs2")

# Subset to clusters of interest: 0 and 1 (confirmed astrocytes), 7 (uncertain)
astro_sub <- subset(astro, seurat_clusters %in% c(0, 1, 7))
```

## Canonical marker genes

### Astrocyte identity and reactive markers

```{r dotplot-identity}
#| fig-width: 10
#| fig-height: 5

DefaultAssay(astro_sub) <- "SCT"

# Astrocyte identity markers
astro_markers <- c("Gfap", "Aqp4", "Aldh1l1", "Slc1a3", "Slc1a2", "Gja1", "S100b", "Ckb", "Aldoc")
astro_markers <- intersect(astro_markers, rownames(astro_sub))

DotPlot(astro_sub, features = astro_markers, group.by = "seurat_clusters") +
  coord_flip() +
  labs(title = "Astrocyte identity markers") +
  theme(axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 12, face = "italic"))
```

```{r dotplot-reactive}
#| fig-width: 8
#| fig-height: 4

# Reactive / inflammatory markers
reactive_markers <- c("Cxcl10", "Ccl2", "Vim", "Serpina3n", "Lcn2", "C3", "Tnf")
reactive_markers <- intersect(reactive_markers, rownames(astro_sub))

DotPlot(astro_sub, features = reactive_markers, group.by = "seurat_clusters") +
  coord_flip() +
  labs(title = "Reactive / inflammatory markers") +
  theme(axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 12, face = "italic"))
```

### Top marker genes per cluster

```{r fiddle-top-markers}
#| column: screen
#| fig-width: 15
#| fig-height: 4

# Read precomputed markers, filter to clusters 0, 1, 7, pick top 5 per cluster
top_genes <- read_csv("results/20260305-astrocyte_all_markers_0.3.csv",
                      show_col_types = FALSE) |>
  filter(cluster %in% c(0, 1, 7)) |>
  group_by(cluster) |>
  arrange(desc(avg_log2FC), .by_group = TRUE) |>
  slice_head(n = 5) |>
  pull(gene) |>
  unique()

fiddle(astro_sub,
       genes_of_interest = top_genes,
       classification_col = "seurat_clusters",
       assay = "SCT")
```

## Proportions and brain region distribution

### Cluster proportions per sample

```{r proportions}
#| fig-width: 10
#| fig-height: 6

prop_res <- scProp(astro_sub,
                   cluster_col   = "seurat_clusters",
                   sample_col    = "sample_ID",
                   treatment_col = "Treatment.Group")

prop_res$counts_plot / prop_res$proportions_plot
```

```{r proportions-table}
#| results: asis
export2md(prop_res$summary_table)
```

### Cell counts by brain region

```{r region-bar}
#| column: page
#| fig-width: 10
#| fig-height: 5
#| label: region-bar


# Cell counts per Region, faceted by cluster, dodged by treatment
astro_sub@meta.data |>
  filter(!is.na(Region)) |>
  dplyr::count(seurat_clusters, Region, Treatment.Group) |>
  ggplot(aes(x = Region, y = n, fill = Treatment.Group)) +
  geom_col(position = "dodge", alpha = 0.85) +
  facet_wrap(~seurat_clusters, scales = "free_y", labeller = label_both) +
  scale_fill_manual(values = c("Adu" = "#d73027", "IgG" = "#4575b4")) +
  labs(title = "Cell counts per brain region",
       y = "Number of cells", x = NULL, fill = NULL) +
  theme_minimal(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank(),
        strip.text = element_text(size = 14, face = "bold"))
```

## Pairwise DE between clusters

Pseudobulk DESeq2 comparing cluster pairs. For each comparison, cells from both
clusters are pooled across all samples, aggregated per sample-cluster combination,
and tested with DESeq2.

```{r pairwise-de-compute}
#| eval: false
#| label: pairwise-de-compute

DefaultAssay(astro) <- "Xenium"

# Define pairwise comparisons
pairs <- list(c("0", "1"), c("0", "7"), c("1", "7"))

pairwise_de <- map(pairs, \(pr) {
  # Subset to the two clusters
  sub_obj <- subset(astro, seurat_clusters %in% pr)

  # Create combined sample-cluster ID for pseudobulk
  sub_obj$sample_cluster <- paste0(sub_obj$sample_ID, "_cl", sub_obj$seurat_clusters)

  # Pseudobulk aggregation per sample-cluster
  bulk <- AggregateExpression(sub_obj, group.by = "sample_cluster", return.seurat = TRUE)

  # Extract cluster label from column names (e.g., KK4-465-cl0 -> 0)
  bulk$cluster_label <- gsub(".*cl", "", colnames(bulk))
  Idents(bulk) <- "cluster_label"

  # DESeq2 via anayet
  anayet(bulk,
         ident1          = pr[1],
         ident2          = pr[2],
         test_method     = "DESeq2",
         min_pct         = 0,
         p_adj_threshold = 0.1,
         plot_title      = paste("Cluster", pr[1], "vs", pr[2]),
         print_plot      = FALSE)

}) |> set_names(paste0("cl", map_chr(pairs, 1), "_vs_cl", map_chr(pairs, 2)))

saveRDS(pairwise_de, "results/20260305-pairwise_DE_astro_clusters.rds")
```

### DEG counts per comparison

The bar chart below tallies the number of significant DEGs (adjusted *p* < 0.1) for each
pairwise cluster comparison, stratified by direction of effect: genes with higher expression
in the first cluster of the pair (red) versus genes with higher expression in the second (blue).

```{r}
#| fig-width: 5
#| fig-height: 4
#| label: pairwise-deg-counts


library(pheatmap)

pairwise_de <- readRDS("results/20260305-pairwise_DE_astro_clusters.rds")

# Bar chart of up/down DEG counts per comparison
deg_counts <- imap(pairwise_de, \(res, comp) {
  if (nrow(res$significant_results) == 0) return(NULL)
  label <- gsub("cl(\\d+)_vs_cl(\\d+)", "\\1 vs \\2", comp)
  res$significant_results |>
    rownames_to_column("gene") |>
    mutate(Comparison = label,
           Direction  = ifelse(avg_log2FC > 0, "Up in first", "Down in first"))
}) |>
  compact() |>
  list_rbind() |>
  dplyr::count(Comparison, Direction)

ggplot(deg_counts, aes(x = Comparison, y = n, fill = Direction)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.9) +
  scale_fill_manual(values = c("Up in first" = "#d73027", "Down in first" = "#4575b4")) +
  labs(title = "Significant DEGs per pairwise comparison",
       subtitle = "padj < 0.1",
       x = NULL, y = "Number of DEGs", fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        axis.text.x = element_text(size = 14))
```

### Top DEG heatmap

To highlight genes that most strongly distinguish the three clusters, a heatmap displays the
log~2~ fold change for the top 20 most significant DEGs (by adjusted *p*-value) from each
pairwise comparison. Each column represents a cluster pair (e.g., "0 vs 1") and each row a
gene. The colour scale is symmetric around zero, with red indicating higher expression in the
first cluster and blue indicating higher expression in the second. Asterisks mark gene–comparison
combinations that individually reach significance (adjusted *p* < 0.1). Genes are
hierarchically clustered by their fold-change profiles to reveal shared and comparison-specific
expression patterns.

```{r}
#| fig-width: 10
#| fig-height: 25
#| label: pairwise-heatmap

# Collect all results into a long table with log2FC and padj per comparison
all_de_long <- imap(pairwise_de, \(res, comp) {
  label <- gsub("cl(\\d+)_vs_cl(\\d+)", "\\1 vs \\2", comp)
  res$all_results |>
    rownames_to_column("gene") |>
    mutate(Comparison = label)
}) |> list_rbind()

# Select top 20 DEGs per comparison by padj
top_genes <- all_de_long |>
  filter(p_val_adj < 0.1) |>
  group_by(Comparison) |>
  arrange(p_val_adj, .by_group = TRUE) |>
  slice_head(n = 20) |>
  pull(gene) |>
  unique()

if (length(top_genes) > 0) {
  # Build log2FC matrix (genes x comparisons)
  fc_matrix <- all_de_long |>
    filter(gene %in% top_genes) |>
    pivot_wider(id_cols = gene, names_from = Comparison, values_from = avg_log2FC) |>
    column_to_rownames("gene") |>
    as.matrix()

  # Build significance matrix for asterisk overlay
  sig_matrix <- all_de_long |>
    filter(gene %in% top_genes) |>
    pivot_wider(id_cols = gene, names_from = Comparison, values_from = p_val_adj) |>
    column_to_rownames("gene") |>
    as.matrix()

  display_matrix <- ifelse(!is.na(sig_matrix) & sig_matrix < 0.1, "*", "")

  clim <- ceiling(max(abs(fc_matrix), na.rm = TRUE))

  pheatmap(fc_matrix,
    color           = colorRampPalette(c("#4575b4", "white", "#d73027"))(100),
    breaks          = seq(-clim, clim, length.out = 101),
    na_col          = "grey88",
    display_numbers = display_matrix,
    fontsize_number = 24,
    number_color    = "black",
    cluster_rows    = TRUE,
    cluster_cols    = FALSE,
    main            = "Top DEGs — pairwise cluster comparisons\n",
    fontsize        = 18,
    fontsize_row    = 13,
    angle_col       = 45,
    cellwidth       = 30,
    cellheight      = 30
  )
} else {
  message("No significant DEGs found across any comparison — heatmap skipped.")
}
```
