---
title: "Astrocyte subpopulations after cleaning"
subtitle: "Identification of Astrocyte subpopulations after subsetting and reclustering"
date: last-modified
format:
  html:
    number-sections: false
---

```{r}
#| label: setup
#| message: false
library(qs2)
library(Seurat)
library(patchwork)
library(tidyverse)
library(kableExtra)
library(DESeq2)
library(gridExtra)


astro <- qs2::qs_read("seurat_objects/20260309-astro_cleaned_0.3.qs2")

```


```{r}
#| eval: false
#| label: SCTransform

# Load seurat with subsetted astrocytes object 
astro_subset <- qs2::qs_read("seurat_objects/20260305-astro_0.3.qs2")

# genes to remove
genes_to_remove <- read_csv("results/20260309_astrocyteSubcluster_genes_removed.csv") |> pull(Gene)
all_features <- Features(astro_subset)
keep_genes <- all_features[!all_features %in% genes_to_remove]


astro_cleaned  <- subset(astro_subset, features = keep_genes)
# Keep only the assay needed for fresh SCTransform
DefaultAssay(astro_cleaned) <- "Xenium"
astro0 <- DietSeurat(astro_cleaned, assays = "Xenium", dimreducs = NULL, graphs = NULL)
# Drop Xenium image/FOV segmentation data; otherwise merge triggers huge future globals
astro0@images <- list()

# SCTransform on all astrocytes together (no split/merge, no integration)
astro <- SCTransform(
  astro0,
  assay = "Xenium",
  new.assay.name = "SCT",
  verbose = FALSE
)
```




## PCA and dimensionality selection

```{r}
#| eval: false
#| label: Run PCA on SCT assay
DefaultAssay(astro) <- "SCT"
astro <- RunPCA(
  astro,
  assay   = "SCT",
  features = VariableFeatures(astro),
  verbose = FALSE)
```
```{r}
#| fig-width: 5
#| fig-height: 3
#| label: Elbow plot
# Elbow plot to visualise variance explained per PC
ElbowPlot(astro, ndims = 50) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red") +
  labs(title = "Elbow plot") +
  theme_minimal()
```

```{r}
#| fig-height: 5
#| label: PCA biplots

# PCA biplots coloured by sample and treatment
p1 <- DimPlot(astro, reduction = "pca", group.by = "Slide") +
  labs(title = "PCA — Slide")
p2 <- DimPlot(astro, reduction = "pca", group.by = "Treatment.Group") +
  labs(title = "PCA — Treatment")
p1 + p2
```

```{r}
#| label: neighbors & UMAP
#| eval: false

# Set seed for reproducibility
set.seed(1234)

# Compute nearest neighbors using first 30 PCs
astro <- FindNeighbors(astro, reduction = "pca", dims = 1:30, graph.name = "astro_snn")

# Perform clustering at multiple resolutions for later comparison
astro <- FindClusters(astro, graph.name = "astro_snn", resolution = 0.3)

# Compute UMAP for visualization
astro <- RunUMAP(astro, reduction = "pca", dims = 1:30, verbose = FALSE)

# qs_save(astro, "seurat_objects/20260309-astro_cleaned_0.3.qs2")
```

resolution = 0.3 

```{r}
#|fig-width: 5
#|fig-height: 5
#| label: plot UMAPs 1
DimPlot(astro)
```

```{r}
#| fig-height: 5
#| label: plot UMAPs 2
DimPlot(astro, split.by = "Treatment.Group")
```

## Top marker genes per cluster

`FindAllMArkers` run on log-normalized residual expression values from SCTransform in the SCT assay (Values are continuous and typically centered around 0)
```{r}
#| eval: false
#| label: find markers
Idents(astro) <- astro$seurat_clusters
DefaultAssay(astro) <- "SCT"
all_markers <- FindAllMarkers(astro, only.pos = TRUE, densify = TRUE)
# write_csv(all_markers, file = "results/20260309-astrocyte_cleaned_all_markers_0.3.csv")
```

```{r}
#| label: top markers per cluster
#| column: screen
#| fig-width: 15
#| fig-height: 4
source("20260217-helper_functions.R")

# Top 5 markers per cluster by avg_log2FC
top_genes <- read_csv("results/20260309-astrocyte_cleaned_all_markers_0.3.csv",
                      show_col_types = FALSE) |>
  group_by(cluster) |>
  arrange(desc(avg_log2FC), .by_group = TRUE) |>
  slice_head(n = 5) |>
  pull(gene) |>
  unique()

fiddle(astro,
       genes_of_interest = top_genes,
       classification_col = "seurat_clusters",
       assay = "SCT")
```

### Cluster identity assessment

Based on the top 5 marker genes per cluster, we assessed whether each cluster represents true astrocytes or contaminating cell types from the initial subsetting.



## Astrocyte markers


```{r}
#| label: astrocyte-markers-featureplot
#| column: page
#| fig-height: 12
#| fig-width: 10

FeaturePlot(astro,features = c("Aldh1l1",	"Aldoc",	"Gfap",	"Slc7a10",	"Aqp4"), ncol=2, min.cutoff = "q20")
```

```{r}
#| label: astrocyte-markers-violin
#| column: page
#| fig-height: 12
#| fig-width: 10

VlnPlot(astro,features = c("Aldh1l1",	"Aldoc",	"Gfap",	"Slc7a10",	"Aqp4"),alpha = 0, ncol=2)
```




