---
title: "Differentially expressed genes in Fibroblast across brain regions"
date: last-modified
---

```{r setup}
#| message: false
library(qs2)
library(Seurat)
library(DESeq2)
library(patchwork)
library(gridExtra)
library(pheatmap)
library(tidyverse)
library(kableExtra)


source("20260217-helper_functions.R")


full <- qs_read("seurat_objects/20260217_Part2_fullobj.qs2")
fibroblast  <- subset(full, annotatedclusters == "Fibroblast")
```

# DE

To characterise region-specific transcriptional responses of Fibroblast cells to aducanumab treatment,
Fibroblast cells were isolated from the full spatial transcriptomics object on the basis of the
`annotatedclusters` annotation.

The Xenium raw counts assay was set as the default to ensure
that downstream pseudobulk aggregation and DESeq2 normalization operated on unmodified integer
counts. Cells lacking a brain-region assignment were excluded prior to differential expression
analysis to avoid confounding by spatially unassigned transcriptomes.

## Cell counts per region

The table below enumerates the total number of Fibroblast cells captured per brain region and
their distribution across treatment groups (aducanumab, Adu; isotype control IgG).

```{r region-summary}
fibroblast@meta.data |>
  group_by(Region) |>
  summarise(
    n_cells = n(),
    n_Adu   = sum(Treatment.Group == "Adu"),
    n_IgG   = sum(Treatment.Group == "IgG")
  ) |>
  arrange(desc(n_cells)) |>
  kbl(caption = "Fibroblast per brain region") |>
  kable_styling("striped")
```

```{r remove-NA-regions}
DefaultAssay(fibroblast) <- "Xenium"
fibroblast <- subset(fibroblast, !is.na(Region))
```

## Pseudobulk DE (Adu vs IgG) per region

To identify transcriptional changes induced by aducanumab in a spatially resolved manner,
pseudobulk differential expression analysis was performed independently for each brain region.

Within each region, raw Xenium counts were summed per biological sample using
`AggregateExpression`, yielding one pseudobulk profile per animal (n = 3 Adu, n = 3 IgG).
Differential expression between treatment groups was then assessed using DESeq2, as implemented
via Seurat's `FindMarkers` interface with `test.use = "DESeq2"` and no minimum expression
filter (`min.pct = 0`), ensuring all detected genes were considered.

For each eligible region, results are displayed as a volcano plot, highlighting genes reaching
statistical significance (Benjamini-Hochberg-adjusted *p* < 0.1), alongside a table of
significant differentially expressed genes (DEGs), ranked by adjusted *p*-value.

```{r pseudobulk-de}
#| results: asis
#| fig.width: 14
#| fig.height: 6
#| warning: false

regions <- sort(unique(fibroblast$Region))

de_by_region <- map(regions, \(reg) {
  sub_obj <- subset(fibroblast, Region == reg)

  # Replication guard: DESeq2 needs >=2 samples per treatment group
  n_per_group <- table(factor(
    distinct(sub_obj@meta.data, sample_ID, Treatment.Group)$Treatment.Group,
    levels = c("Adu", "IgG")
  ))
  if (any(n_per_group < 2)) {
    cat("\n\n###", reg, "\n\n_Skipped: fewer than 2 samples per treatment group._\n\n")
    return(NULL)
  }

  bulk <- AggregateExpression(sub_obj, group.by = "sample_ID", return.seurat = TRUE)

  # AggregateExpression converts _ to - (e.g. KK4_465 -> KK4-465)
  sample_meta <- distinct(sub_obj@meta.data, sample_ID, Treatment.Group) |>
    mutate(bulk_id = gsub("_", "-", sample_ID))
  bulk$Treatment.Group <- sample_meta$Treatment.Group[
    match(colnames(bulk), sample_meta$bulk_id)
  ]
  Idents(bulk) <- "Treatment.Group"

  res <- anayet(bulk,
                ident1          = "Adu",
                ident2          = "IgG",
                test_method     = "DESeq2",
                min_pct         = 0,
                p_adj_threshold = 0.1,
                plot_title      = paste("Fibroblast —", reg),
                print_plot      = FALSE,
                min_cells_group = 2)

  cat("\n\n###", reg, "\n\n")

  sig_tbl <- if (nrow(res$significant_results) > 0) {
    res$significant_results |>
      select(p_val, avg_log2FC, p_val_adj) |>
      mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
      arrange(p_val_adj) |>
      tableGrob(theme = ttheme_minimal(base_size = 18))
  } else {
    tableGrob(
      data.frame(Result = "No significant DEGs"),
      theme = ttheme_minimal(base_size = 20)
    )
  }

  print(res$volcano_plot + wrap_elements(sig_tbl) + plot_layout(widths = c(2, 1)))
  res

}) |> set_names(regions) |> compact()
```

```{r save-de-results}

saveRDS(
  map(de_by_region, \(res) res$all_results),
  "results/20260310-DEG_Fibroblast_byRegion.rds"
)
```

```{r save-degs}
# Build one-row CSV summarising DEGs per region (gene names + counts)
if (length(de_by_region) > 0) {

  deg_summary <- imap(de_by_region, \(res, reg) {
    sig <- res$significant_results |> rownames_to_column("gene")
    up_genes   <- sig |> filter(avg_log2FC > 0) |> pull(gene)
    down_genes <- sig |> filter(avg_log2FC < 0) |> pull(gene)
    tibble(
      region = reg,
      up     = paste(up_genes, collapse = ", "),
      down   = paste(down_genes, collapse = ", "),
      total  = nrow(sig)
    )
  }) |> list_rbind()

  deg_wide <- deg_summary |>
    pivot_wider(names_from = region,
                values_from = c(up, down, total),
                names_glue = "{region}_{.value}") |>
    mutate(cell_type = "Fibroblast", .before = 1)
} else {
  message("No regions with sufficient replication — writing empty summary.")
  deg_wide <- tibble(cell_type = "Fibroblast")
}

if (!dir.exists("deg_number")) dir.create("deg_number")
write_csv(deg_wide, "deg_number/DEG_Fibroblast.csv")
```

## Summary visualisations

The following figures integrate findings across all brain regions to facilitate cross-regional
comparison of Fibroblast transcriptional responses to aducanumab treatment.

### DEG counts per region

To provide an overview of the magnitude and directionality of transcriptional responses across
brain regions, the total number of significant DEGs (Benjamini-Hochberg-adjusted *p* < 0.1)
was tallied per region and stratified by direction of effect: genes upregulated in aducanumab-
treated animals relative to IgG controls (Up in Adu) and genes downregulated (Down in Adu).
Regions with no significant DEGs are absent from the plot.

```{r deg-counts}
#| fig.width: 4
#| fig.height: 5

deg_counts <- imap(de_by_region, \(res, reg) {
  if (nrow(res$significant_results) == 0) return(NULL)
  res$significant_results |>
    rownames_to_column("gene") |>
    mutate(Region    = reg,
           Direction = ifelse(avg_log2FC > 0, "Up in Adu", "Down in Adu"))
}) |>
  compact() |>
  list_rbind()

if (nrow(deg_counts) == 0) {
  message("No significant DEGs in any region — bar chart skipped.")
} else {
deg_counts <- count(deg_counts, Region, Direction)
ggplot(deg_counts, aes(x = Region, y = n, fill = Direction)) +
  geom_col(position = "dodge", width = 0.7, alpha = 0.9) +
  scale_fill_manual(values = c("Up in Adu" = "#d73027", "Down in Adu" = "#4575b4")) +
  labs(
    title    = "Significant DEGs per brain region",
    subtitle = "padj < 0.1",
    x        = NULL,
    y        = "Number of DEGs",
    fill     = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
  axis.text.x = element_text(size = 16,angle = 45, hjust = 1))
}
```

### Cross-region log2FC heatmap

To identify genes with consistent or divergent regulation across brain regions, a hierarchically
clustered heatmap was constructed for all genes reaching statistical significance (adjusted
*p* < 0.1) in at least one region.

Each cell displays the log~2~ fold change (Adu vs IgG) for
that gene–region combination. The colour scale is symmetric and centred at zero, with red
denoting upregulation in aducanumab-treated animals and blue denoting downregulation. Asterisks
mark gene–region combinations that individually meet the adjusted *p* < 0.1 threshold. Genes and regions
were clustered hierarchically by their fold-change profiles to reveal co-regulated gene modules and similarities between brain regions.

```{r heatmap}
#| fig.width: 10
#| fig.height: 10



all_de_long <- imap(de_by_region, \(res, reg) {
  res$all_results |> rownames_to_column("gene") |> mutate(Region = reg)
}) |> list_rbind()

sig_genes <- all_de_long |>
  filter(p_val_adj < 0.1) |>
  pull(gene) |>
  unique()

if (length(sig_genes) > 0) {
  fc_matrix <- all_de_long |>
    filter(gene %in% sig_genes) |>
    pivot_wider(id_cols = gene, names_from = Region, values_from = avg_log2FC) |>
    column_to_rownames("gene") |>
    as.matrix()

  sig_matrix <- all_de_long |>
    filter(gene %in% sig_genes) |>
    pivot_wider(id_cols = gene, names_from = Region, values_from = p_val_adj) |>
    column_to_rownames("gene") |>
    as.matrix()

  display_matrix <- ifelse(!is.na(sig_matrix) & sig_matrix < 0.1, "*", "")

  clim <- ceiling(max(abs(fc_matrix), na.rm = TRUE))

  pheatmap(fc_matrix,
    color         = colorRampPalette(c("#4575b4", "white", "#d73027"))(100),
    breaks        = seq(-clim, clim, length.out = 101),
    na_col        = "grey88",
    display_numbers = display_matrix,
    fontsize_number = 24,
    number_color  = "black",
    cluster_rows  = F,
    cluster_cols  = T,
    main          = "Fibroblast log2FC (Adu vs IgG) by region\n",
    fontsize      = 18,
    fontsize_row  = 15,
    angle_col     = 45,
    cellwidth = 30,
    cellheight = 30
  )
} else {
  message("No significant DEGs found across any region — heatmap skipped.")
}
```

### Effect size dot plot

To simultaneously visualise the effect size, statistical significance, and regional distribution
of all significant DEGs, a dot plot was generated in which each point represents a single
gene–region observation.

The horizontal position encodes the log~2~ fold change (Adu vs IgG),
dot size reflects the strength of statistical evidence (−log~10~ adjusted *p*-value), and colour
denotes the brain region. Genes are ordered along the vertical axis by their mean log~2~ fold
change across all regions in which they were significant, placing consistently downregulated
genes at the top and upregulated genes at the bottom.

```{r dotplot}
#| fig.width: 9
#| fig.height: 3

sig_all <- imap(de_by_region, \(res, reg) {
  res$significant_results |> rownames_to_column("gene") |> mutate(Region = reg)
}) |> list_rbind()

if (nrow(sig_all) > 0) {
  gene_order <- sig_all |>
    group_by(gene) |>
    summarise(mfc = mean(avg_log2FC)) |>
    arrange(mfc) |>
    pull(gene)

  sig_all |>
    mutate(gene = factor(gene, levels = gene_order)) |>
    ggplot(aes(x = avg_log2FC, y = gene,
               colour = Region,
               size   = -log10(p_val_adj + 1e-300))) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_point(alpha = 0.85) +
    scale_size_continuous(
      name  = expression(-log[10](p[adj])),
      range = c(2, 8)
    ) +
    scale_colour_brewer(palette = "Set2") +
    labs(
      title = "Fibroblast DEGs — all brain regions\n",
      x     = expression(log[2] * "FC  (Adu / IgG)"),
      y     = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      axis.text         = element_text(size= 16,face = "italic"),
      axis.title = element_text(size = 18),
      plot.title = element_text(size = 20),
      axis.ticks = element_blank(),
      panel.grid.major.x  = element_blank()
    )
} else {
  message("No significant DEGs found — dot plot skipped.")
}
```


