---
title: "DEG summary across cell types and brain regions"
date: last-modified
format:
  html:
    number-sections: false
---

This report consolidates differential expression results across cell
types and five brain regions
(Cortex, Hippocampus, Hypothalamus, Thalamus, White Matter).


**DE testing.** For each cell type and brain region, cells were pseudobulk-aggregated per
sample using `AggregateExpression()` on the Xenium raw-counts assay (integer counts, no
normalisation). Differential expression was then tested with DESeq2 via Seurat's
`FindMarkers(test.use = "DESeq2")`, comparing Adu vs IgG. Regions with fewer than two
biological replicates per group were excluded. Genes with adjusted *p*-value < 0.1 were
considered significant.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(pheatmap)
library(patchwork)
library(UpSetR)

# auto-discover all per-cell-type DEG CSVs
csv_files <- list.files("deg_number", pattern = "\\.csv$", full.names = TRUE)
deg_wide  <- map(csv_files, read_csv, show_col_types = FALSE) |> list_rbind()

# extract region names from the _total columns
regions <- str_remove(names(deg_wide)[str_detect(names(deg_wide), "_total$")], "_total")

# pivot total counts to long format
deg_total <- deg_wide |>
  select(cell_type, ends_with("_total")) |>
  pivot_longer(-cell_type, names_to = "region", values_to = "total") |>
  mutate(region = str_remove(region, "_total"))

# pivot up/down counts to long format
deg_long <- deg_wide |>
  select(cell_type, ends_with("_up"), ends_with("_down")) |>
  pivot_longer(-cell_type, names_to = "col", values_to = "genes") |>
  mutate(
    direction = if_else(str_detect(col, "_up$"), "Up", "Down"),
    region    = str_remove(col, "_(up|down)$")
  ) |>
  select(-col)

# count genes per cell_type/region/direction
deg_counts <- deg_long |>
  mutate(count = map_int(genes, \(x) {
    if (is.na(x) || x == "") 0L else length(str_split_1(x, ",\\s*"))
  })) |>
  select(cell_type, region, direction, count)
```

# Total DEGs per cell type and region

Hierarchically clustered heatmap of total DEG counts (**padj < 0.1**) across all cell types
and brain regions. Rows and columns are clustered to highlight similarities. Numbers inside
cells indicate the count of significant DEGs.

```{r}
#| label: heatmap-total
#| fig.width: 8
#| column: page
#| fig.height: 12

# build matrix: rows = cell types, cols = regions; sort rows alphabetically
mat <- deg_total |>
  pivot_wider(names_from = region, values_from = total) |>
  column_to_rownames("cell_type") |>
  as.matrix()
mat <- mat[sort(rownames(mat)), ]

num_mat <- matrix(ifelse(is.na(mat), "", sprintf("%.0f", mat)), nrow = nrow(mat), dimnames = dimnames(mat))

pheatmap(mat,
  color           = colorRampPalette(c("white", "#d73027"))(100),
  display_numbers = num_mat,
  fontsize_number = 14,
  number_color    = "black",
  cluster_rows    = F,
  cluster_cols    = TRUE,
  main            = "Total DEGs by cell type and brain region\n",
  fontsize        = 14,
  fontsize_row    = 12,
  fontsize_col    = 14,
  angle_col       = 45,
  cellwidth       = 50,
  cellheight      = 30
)
```

# Up vs Down DEGs

Dodged bar chart showing the number of upregulated and downregulated DEGs for each cell type,
faceted by brain region. This allows comparison of directionality across cell types within each
region.

```{r}
#| label: bar-chart-direction
#| column: page
#| fig.width: 12
#| fig.height: 6

deg_counts |>
  ggplot(aes(x = cell_type, y = count, fill = direction)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~ region, nrow = 1) +
  scale_fill_manual(values = c(Up = "#d73027", Down = "#4575b4")) +
  labs(
    title = "DEGs by direction, cell type, and brain region\n",
    x     = NULL,
    y     = "Number of DEGs",
    fill  = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x      = element_text(size = 11, angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    strip.text       = element_text(size = 13, face = "bold")
  )
```

# Up and Down heatmaps

Side-by-side heatmaps of upregulated and downregulated DEG counts, allowing comparison
of directionality patterns across cell types and regions.

```{r}
#| label: heatmap-updown
#| fig.width: 14
#| fig.height: 12


# build separate matrices for up and down; sort rows alphabetically
mat_up <- deg_counts |>
  filter(direction == "Up") |>
  pivot_wider(names_from = region, values_from = count, id_cols = cell_type) |>
  column_to_rownames("cell_type") |>
  as.matrix()
mat_up <- mat_up[sort(rownames(mat_up)), ]

mat_down <- deg_counts |>
  filter(direction == "Down") |>
  pivot_wider(names_from = region, values_from = count, id_cols = cell_type) |>
  column_to_rownames("cell_type") |>
  as.matrix()
mat_down <- mat_down[sort(rownames(mat_down)), ]

# shared color ceiling
max_val <- max(c(mat_up, mat_down), na.rm = TRUE)

# draw both heatmaps
p_up <- pheatmap(mat_up,
  color           = colorRampPalette(c("white", "#d73027"))(100),
  breaks          = seq(0, max_val, length.out = 101),
  display_numbers = TRUE,
  number_format   = "%.0f",
  fontsize_number = 14,
  number_color    = "black",
  cluster_rows    = FALSE,
  cluster_cols    = FALSE,
  main            = "Upregulated DEGs\n",
  fontsize        = 14,
  fontsize_row    = 12,
  fontsize_col    = 14,
  angle_col       = 45,
  cellwidth       = 50,
  cellheight      = 30,
  silent          = TRUE
)

p_down <- pheatmap(mat_down,
  color           = colorRampPalette(c("white", "#4575b4"))(100),
  breaks          = seq(0, max_val, length.out = 101),
  display_numbers = TRUE,
  number_format   = "%.0f",
  fontsize_number = 14,
  number_color    = "black",
  cluster_rows    = FALSE,
  cluster_cols    = FALSE,
  main            = "Downregulated DEGs\n",
  fontsize        = 14,
  fontsize_row    = 12,
  fontsize_col    = 14,
  angle_col       = 45,
  cellwidth       = 50,
  cellheight      = 30,
  silent          = TRUE
)

# arrange side by side
gridExtra::grid.arrange(p_up$gtable, p_down$gtable, ncol = 2)
```



# Gene overlap across cell types

For each brain region, identify DEGs shared across multiple cell types. Genes appearing in
more than one cell type within the same region may indicate conserved responses to treatment.

```{r}
#| label: gene-overlap
#| fig.width: 12
#| fig.height: 8
#| column: page

# parse gene lists into a tidy data frame
gene_df <- deg_long |>
  filter(!is.na(genes) & genes != "") |>
  mutate(gene = str_split(genes, ",\\s*")) |>
  unnest(gene) |>
  select(cell_type, region, direction, gene)

# for each region, count how many cell types share each gene
shared_genes <- gene_df |>
  distinct(cell_type, region, gene) |>
  group_by(region, gene) |>
  summarise(
    n_cell_types = n(),
    cell_types   = paste(cell_type, collapse = ", "),
    .groups      = "drop"
  ) |>
  filter(n_cell_types > 1) |>
  arrange(region, desc(n_cell_types))

if (nrow(shared_genes) > 0) {
  shared_genes |>
    kableExtra::kbl(
      caption = "DEGs shared across cell types within each brain region",
      col.names = c("Region", "Gene", "# Cell types", "Cell types")
    ) |>
    kableExtra::kable_styling("striped", full_width = FALSE)
} else {
  cat("No DEGs shared across multiple cell types in any region.")
}
```

```{r}
#| label: upset-plot
#| fig.width: 10
#| fig.height: 10
#| column: page
# UpSet plot: for each region with shared genes, show cell-type overlaps
# pool all genes across regions for global cell-type overlap view
gene_by_ct <- gene_df |>
  distinct(cell_type, gene) |>
  mutate(present = 1) |>
  pivot_wider(names_from = cell_type, values_from = present, values_fill = 0) |>
  as.data.frame()

if (nrow(gene_by_ct) > 1 && ncol(gene_by_ct) > 2) {
  upset(gene_by_ct,
    sets      = setdiff(names(gene_by_ct), "gene"),
    order.by  = "freq",
    text.scale = 1.3,
    mainbar.y.label = "Shared DEGs",
    sets.x.label    = "DEGs per cell type"
  )
} else {
  cat("Not enough overlap data for UpSet plot.")
}
```
